---
title: 'Data cleaning: Unemployment'
author: "Gabriel Urbina"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
# Setup: ----

# path to files: ----

# path to data
data <- "/media/gabriel/GABOMOVIL/AGUN_personal_projects/Data_visualization_case_study_Barcelona/Data/"

# path to documents

documents <- "/media/gabriel/GABOMOVIL/AGUN_personal_projects/Data_visualization_case_study_Barcelona/Documents/"

# path to graphs

graphs <- "/media/gabriel/GABOMOVIL/AGUN_personal_projects/Data_visualization_case_study_Barcelona/Outcomes/Graphs/"

# path to tables

tables <- "/media/gabriel/GABOMOVIL/AGUN_personal_projects/Data_visualization_case_study_Barcelona/Outcomes/Tables/"

# knitr setup: ----
knitr::opts_chunk$set(
  dpi = 400,
  results = "asis",
  message= F,
  warning = F,
  echo = TRUE)

# load packages: ----
pacman::p_load(
  "data.table",
  "tidyverse",
  "kable",
  "kableExtra",
  "psych",
  "readr",
  "lubridate",
  "hms"
)


```

<!-- Data exploration --->
# What do I have in the data set "unemployment"? {-}

First, I loaded the data set. Then I explored each variable and drew a first road map of the next steps. First,  I renamed some variables because their names are formed by two words separated by blank space. I replaced the blank space by "_". The data set contains data from 2013 to 2017. We want only 2017, so I kept only data for year equal to 2017. "NA" values were present and coded as "No consta". I renamed "No consta" to "NA" values. The data contained number of unemployed people per gender. I created a unique value per district-neighborhood-month. In consequence, I removed gender. As last step, I kept only the variables of interest. The last step consists on detecting duplicated cases. The result showed that there were no duplicated cases. 

From the 888 observation, only 870 has no one NA on any of its rows. There are 18 cases for which district code, district name, neighborhood code & neighborhood name are all NA. Therefore, I removed the rows with NA. 

```{r}

# load data set: unemployment ----

unemployment <- fread(paste0(data, "unemployment.csv"))

# get variable names: 

names(unemployment)

# modify variables names: ----

colnames(unemployment) <- 
  gsub(" ", 
       "_",
       colnames(unemployment))

# get a summary of the data set: ----

summary(unemployment)

# keep only data for 2017: ----

unemployment.2017 <- 
  unemployment[Year == "2017"]

# summary: unemployment 2017 ----

summary(unemployment.2017)
# set NA values: characters ----

unemployment.2017 <- unemployment.2017[, 
                  lapply(.SD, function(x) ifelse(
                  x == "No consta", NA, x
                  )), .SDcols = names(unemployment.2017)]

# set NA values: integers ----

unemployment.2017 <- unemployment.2017[,
                    lapply(
                          .SD,
                          function(x) ifelse(
                            x == 99,
                            NA,
                            x
                          )), 
                    .SDcols = names(
                      unemployment.2017
                    )]

# create the total unemployment column: ----
# at neighborhood frequency

unemployment.2017 <-
  unemployment.2017[,
               unemployment :=
                 sum(Number, 
                     na.rm = TRUE),
               by = c(               
                 "District_Name",
                 "Neighborhood_Name"
               )
]

# generate unemployment by sex: ----

unemployment.2017 <-
  unemployment.2017[,
   c("female_unemployment",
     "male_unemployment"):=
     .(max(ifelse(
       Gender == "Female",
       sum(Number,
         na.rm = TRUE),
       0)),
       max(ifelse(
         Gender == "Male",
         sum(Number,
           na.rm = TRUE),
         0))),
   by = c(
     "District_Name",
     "Neighborhood_Name",
     "Gender"
   )
  ]

# replace zeros by the maximum value per neighborhood: ----
# values per sex are side by side

unemployment.2017 <-
  unemployment.2017[,
   c("female_unemployment",
     "male_unemployment"):=
     .(max(female_unemployment),
       max(male_unemployment)),
   by =c(
     "District_Name",
     "Neighborhood_Name"
   )]

# create an index variable: Month-District_Name,
# Neighborhood_Name ----

unemployment.2017 <-
  unemployment.2017[, index := 1:.N,
                   by= c(
# at neighboorhood level
      "Neighborhood_Code")
]

# keep one observation per Month-District-Neighborhood: ----

unemployment.2017 <-
  unemployment.2017[index == 1]

# keep variables of interest: ----

unemployment.2017 <- unemployment.2017[,
                  .(District_Code,
                    District_Name,
                    Neighborhood_Code,
                    Neighborhood_Name, 
                    unemployment,
                    female_unemployment,
                    male_unemployment)]

# how many duplicated cases exist ----

table(duplicated(unemployment.2017))

# check duplicated neighborhood names per district ----
table(
  unemployment.2017$District_Name,
  unemployment.2017$Neighborhood_Name)
# No duplicated cases. 
```

```{r}
# checking how many NA still remains ----

table(complete.cases(unemployment.2017))

# Drop NAs ----

unemployment.2017 <- na.omit(unemployment.2017)


```


# Visualization

There is no visible trend in the data. The median value is similar
during the year. However, there is visible difference in the dispercion 
of the data. 
```{r, eval=F,include=FALSE}
# add level for the month variable
level <- c(
  "January",
  "February",
  "March",
  "April",
  "May",
  "June",
  "July",
  "August",
  "September",
  "October",
  "November",
  "December"
)

# parse Month into factor: ----
unemployment.2017 <-
  unemployment.2017[, 
                    Month := parse_factor(Month, levels = 
                                            level)]

# boxplot unemployment per month ----
plot((unemployment.2017$Month), unemployment.2017$unemployment)
```

In general, there are neighborhood with persistent unemployment. 

```{r}
unemployment.2017 %>% 
  ggplot(aes(as.factor(unemployment.2017$Neighborhood_Name),
     unemployment.2017$unemployment)
     ) +
  geom_point() 


```





